---
title: "analysis"
author: "Calli Quire"
date: "2025-02-14"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Set up
## Download, install, run relevant packages
```{r, warning=FALSE, message=FALSE, echo=FALSE}
# Install relevant packages
install.packages("dplyr")
install.packages("survey")
install.packages("haven")
install.packages("tidyr")
install.packages("ggplot2")
install.packages("ggbeeswarm")
install.packages("tidyverse")
install.packages("hrbrthemes")
install.packages("viridis")
```

```{r message=FALSE, warning=FALSE, include=FALSE}
# Load relevant packages
# + message = FALSE, warning=FALSE
library(dplyr)
library(survey)
library(haven)
library(workflowr)
library(tidyr)
library(ggplot2)
library(ggbeeswarm)
library(tidyverse)
library(hrbrthemes)
library(viridis)
#'
options(survey.lonely.psu='adjust')

# Display Version Information
cat("R package versions:\n")
for (p in c("base", "survey","dplyr"))
{ cat(p, ": ", as.character(packageVersion(p)), "\n")}
```


# Import data
## Download and read data
### 1. Demographic data
```{r}
## Demographic (DEMO) data
## isolate participant identification/sequence number, gender, age, variances, race/Hispanic origin

# 2007-2008
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2007/DataFiles/DEMO_E.XPT", tf <- tempfile(), mode="wb")
DEMO_E <- foreign::read.xport(tf)[,c("SEQN","RIAGENDR","RIDAGEYR","SDMVSTRA","SDMVPSU","WTMEC2YR", "RIDRETH1")]

# 2009-2010
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2009/DataFiles/DEMO_F.xpt", tf <- tempfile(), mode="wb")
DEMO_F <- foreign::read.xport(tf)[,c("SEQN","RIAGENDR","RIDAGEYR","SDMVSTRA","SDMVPSU","WTMEC2YR", "RIDRETH1")]

# 2013-2014
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2013/DataFiles/DEMO_H.XPT", tf <- tempfile(), mode="wb")
DEMO_H <- foreign::read.xport(tf)[,c("SEQN","RIAGENDR","RIDAGEYR","SDMVSTRA","SDMVPSU","WTMEC2YR", "RIDRETH1")]

# 2017-2018
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/DEMO_J.XPT", tf <- tempfile(), mode="wb")
DEMO_J <- foreign::read.xport(tf)[,c("SEQN","RIAGENDR","RIDAGEYR","SDMVSTRA","SDMVPSU","WTMEC2YR", "RIDRETH1")]

```


### 2. Reproductive health question (RHQ) data
```{r}
# Reproductive Health Questions (RHQ)
# isolate sequence number, age at first live birth, has had regular periods in the last 12 mo, reason for not having regular periods in last 12 mo

# 2007-2008
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2007/DataFiles/RHQ_E.XPT", tf <- tempfile(), mode="wb")
RHQ_E <- foreign::read.xport(tf)[,c("SEQN","RHD180", "RHQ031", "RHD042")]

# 2009-2010
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2009/DataFiles/RHQ_F.xpt", tf <- tempfile(), mode="wb")
RHQ_F <- foreign::read.xport(tf)[,c("SEQN","RHD180", "RHQ031", "RHD042")]

# 2013-2014
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2013/DataFiles/RHQ_H.xpt", tf <- tempfile(), mode="wb")
RHQ_H <- foreign::read.xport(tf)[,c("SEQN","RHD180", "RHQ031", "RHD043")]

# 2017-2018
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/RHQ_J.xpt", tf <- tempfile(), mode="wb")
RHQ_J <- foreign::read.xport(tf)[,c("SEQN","RHD180", "RHQ031", "RHD043")]

```


### 3. Bone mineral denisty (BMD) data
```{r}
# Dual-Energy X-Ray Absorptiometry - Total Femur
# isolate sequence number, total femoral BMD

# 2007-2008
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2007/DataFiles/DXXFEM_E.xpt", tf <- tempfile(), mode="wb")
DXX_E <- foreign::read.xport(tf)[,c("SEQN", "DXXNKBMD")]

# 2009-2010
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2009/DataFiles/DXXFEM_F.xpt", tf <- tempfile(), mode="wb")
DXX_F <- foreign::read.xport(tf)[,c("SEQN", "DXXNKBMD")]

# 2013-2014
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2013/DataFiles/DXXFEM_H.xpt", tf <- tempfile(), mode="wb")
DXX_H <- foreign::read.xport(tf)[,c("SEQN", "DXXNKBMD")]

# 2017-2018
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/DXXFEM_J.xpt", tf <- tempfile(), mode="wb")
DXX_J <- foreign::read.xport(tf)[,c("SEQN", "DXXNKBMD")]

```


# Clean
## Combine the data sets
```{r}
# combine data sets for DEMO, RHQ, and DXX into three sub-sets
DEMO <- bind_rows(DEMO_E, DEMO_F, DEMO_H, DEMO_J)
RHQ <- bind_rows(RHQ_E, RHQ_F, RHQ_H, RHQ_J) 
DXX <- bind_rows(DXX_E, DXX_F, DXX_H, DXX_J)
```

## Merge subsets into one main data set
```{r}
# merge DEMO + RHQ
predata <- left_join(DEMO, RHQ, by = "SEQN")

# merge 'predata' = DEMO, RHQ, + DXX
data <- left_join(predata, DXX, by = "SEQN")
```

## Combine any single variables with 2+ codes into single columns 
```{r}
# combine any variables that have two different variable names (e.g., RHD042 = RHD043) into single columns

# combine RHD042, RHD043
data$RHQ04X <- coalesce(data$RHD042, data$RHD043)
```

## Simplify variable/column names
```{r}
data <-
   rename(data, 
     BMD = DXXNKBMD, AFB = RHD180, MENS = RHQ031,
     GENDR = RIAGENDR, AGE = RIDAGEYR, RETH = RIDRETH1, NMENS = RHQ04X
     )
```


## Omit males from data
```{r}
# remove male study participants from analysis data; responses to RIAGENDR coded 1
data <- subset(data, !(GENDR %in% c(1)))
```

## Omit any participants without BMD (DXXBMD), AFB (RHX180), period/year (RHQ03X) data
```{r}
# missing BMD data
data <- drop_na(data, "BMD")

# missing AFB data
data <- drop_na(data, "AFB")

# missing period/year data
data <- drop_na(data, "MENS")
```

## Omit any participants who answered "don't know" (coded 999), or "refused" (coded as 777) to AFB
```{r}
data <- subset(data, !(AFB %in% c(777, 999)))
```

## Omit any participants who answered "don't know" (coded 9), or "refused" (coded as 7) to RHQ03X (Period in last 12 mo?)
```{r}
# removes participants who don't know if they have had a period in the last 12 months
data <- subset(data, !(MENS %in% c(7, 9)))
```

## Omit any participants who answered 5-9, 77, 99 to RHQ04X (Why no period/year?)
```{r}
# removes participants who aren't having periods b/c: periods usually irregular (coded 4), going/gone through menopause (coded 5 or 7), medical conditions/treatments (coded 6 or 8), other (coded 9), refused (coded 77), don't know (coded 99)
# there are multiple codes for some conditions due to coding inconsistencies between sampling years
# excludes data on individuals who may not be in 'good health'-- defined 'good health' in exclusion flowchart
data <- subset(data, !(NMENS %in% c(5, 6, 7, 8, 9, 77, 99)))
```

## Omit any respondents 50+yo
```{r}
# omit participants who answer >/= 50 to RIDAGEYR
# removes any respondent equal or older than 50yo, which is above the average age of menopause onset
# ensures that data is constricted to those who are not in menopause

data <- subset(data, !(AGE %in% c(50, 51, 52, 53, 54, 55, 56, 57, 58, 59, 60, 61, 62, 63, 64, 65, 66, 67, 68, 69, 70, 71, 72, 73, 74, 75, 76, 77, 78, 79, 80)))
```

## Add column for AFB b4 20yo vs at/after 20yo
```{r}
# adds a column for if participants' AFB was *before* age of peak bone mineral density accrual (PBMDA) or *at/after* age of PBMDA
# if birth occurred before participant was 20 yo (accepted age of peak bone mineral density accrual), coded "yes"; if occurred when or after participant was 20 yo, coded "no"
data <- data %>%
  mutate(PkBMD = ifelse(AFB < 20, "< 20", ">/= 20"))
```

## Replace race/ethnicity codes with definitions
```{r}
data <- data %>% mutate(RETH=recode(RETH, '1'='Mexican American', '2'='Other Hispanic', '3' = 'Non-Hispanic White', '4' = 'Non-Hispanic Black', '5' = 'Other'))
```

## Create simplified data frame for analyses
```{r}
# drop all columns that are not the combined columns, age, gender, race/origin or sequence number
# makes simplified data frame (named "df") for analyses purposes
# does not delete the full data frame (named "data")
df <- subset(data, select = -c(SDMVSTRA, SDMVPSU, WTMEC2YR, RHD042, RHD043))
```


# Analysis

### Linear regression: BMD vs AFB 
```{r}
# model 1
m_1 <-
  lm(BMD ~ AFB, data = df)

# view summary
summary(m_1)



# visualize
# scatter plot 
lm_1 <- 
  ggplot(df, aes(x = AFB, y = BMD, col = RETH)) + 
  geom_point() + ggtitle("Bone Mineral Density vs Age at First Birth") + ylab("Femoral Neck BMD (g/cm^2)") + xlab("Age at First Birth (yrs)") + scale_color_viridis("Race/Ethnicity", discrete = TRUE, option = "A")

# view
print(lm_1)

# scatter plot w/ mean range (across all race/ethnicities)
lm_1a <- 
  ggplot(df, aes(x = AFB, y = BMD, col = RETH)) + 
  geom_point() + ggtitle("Bone Mineral Density vs Age at First Birth") + ylab("Femoral Neck BMD (g/cm^2)") + xlab("Age at First Birth (yrs)") + scale_color_viridis("Race/Ethnicity", discrete = TRUE, option = "A") + annotate("rect", xmin = -Inf, xmax = Inf, ymin = 0.7, ymax = 1,
           alpha = .3,fill = "lightblue")

# view
print(lm_1a)

```

### Linear regression: BMD vs PkBMD
```{r}
# model 2
m_2 <-
  lm(BMD ~ PkBMD, data = df)

# view summary
summary(m_2)



# visualize
# violin plot
lm_2 <-
  ggplot(df, aes(x = PkBMD, y = BMD, fill = PkBMD)) +
    geom_violin(alpha = 0.3) +
    geom_jitter(aes(color = RETH), size = 0.4, alpha = 0.6) +
    ggtitle("Bone Mineral Denisty vs Age at First Birth Before & At/After 20yo") + xlab("Age at First Birth (yrs)") + ylab("Femoral Neck BMD (g/cm^2)") + scale_color_viridis("Race/Ethnicity", discrete = TRUE, option = "A") + scale_fill_viridis("Age at First Birth (yrs)", discrete = TRUE, option = "G")

# view the plot
print(lm_2)

# violin plot w/ mean range
lm_2a <-
  ggplot(df, aes(x = PkBMD, y = BMD, fill = PkBMD)) +
    geom_violin(alpha = 0.3) +
    geom_jitter(aes(color = RETH), size = 0.4, alpha = 0.6) +
    ggtitle("Bone Mineral Denisty vs Age at First Birth Before & At/After 20yo") + xlab("Age at First Birth (yrs)") + ylab("Femoral Neck BMD (g/cm^2)") + scale_color_viridis("Race/Ethnicity", discrete = TRUE, option = "A") + scale_fill_viridis("Age at First Birth (yrs)", discrete = TRUE, option = "G") + annotate("rect", xmin = -Inf, xmax = Inf, ymin = 0.7, ymax = 1,
           alpha = .3,fill = "lightblue")

# view the plot
print(lm_2a)


# box plot
lm_3 <-
  ggplot(df, aes(x = PkBMD, y = BMD, fill = PkBMD)) +
    geom_boxplot(alpha = 0.3) +
    geom_jitter(aes(color = RETH), size = 0.4, alpha = 0.6) +
    ggtitle("Bone Mineral Denisty vs Age at First Birth Before & At/After 20yo") + xlab("Age at First Birth (yrs)") + ylab("Femoral Neck BMD (g/cm^2)") + scale_color_viridis("Race/Ethnicity", discrete = TRUE, option = "A") + scale_fill_viridis("Age at First Birth (yrs)", discrete = TRUE, option = "G")

# view the plot
print(lm_3)

# box plot w/ mean range
lm_3a <-
  ggplot(df, aes(x = PkBMD, y = BMD, fill = PkBMD)) +
    geom_boxplot(alpha = 0.3) +
    geom_jitter(aes(color = RETH), size = 0.4, alpha = 0.6) +
    ggtitle("Bone Mineral Denisty vs Age at First Birth Before & At/After 20yo") + xlab("Age at First Birth (yrs)") + ylab("Femoral Neck BMD (g/cm^2)") + scale_color_viridis("Race/Ethnicity", discrete = TRUE, option = "A") + scale_fill_viridis("Age at First Birth (yrs)", discrete = TRUE, option = "G") + annotate("rect", xmin = -Inf, xmax = Inf, ymin = 0.7, ymax = 1,
           alpha = .3,fill = "lightblue")

# view the plot
print(lm_3a)
```

### Linear regression: BMD vs PkBMD by Race/Ethnicity
```{r}
# based on the same LM, just split up by race/ethnicity

# violin plot
lm_4 <-
  ggplot(df, aes(x = PkBMD, y = BMD, fill = RETH)) +
    geom_violin(alpha = 0.3) +
    geom_jitter(size = 0.4, alpha = 0.6) +
    ggtitle("Bone Mineral Denisty vs Age at First Birth Before & At/After 20yo") + xlab("Age at First Birth (yrs)") + ylab("Femoral Neck BMD (g/cm^2)") + scale_fill_viridis("Race/Ethnicity", discrete = TRUE, option = "A")

# view the plot
print(lm_4)


# box plot
lm_5 <-
  ggplot(df, aes(x = PkBMD, y = BMD, fill = RETH)) +
    geom_boxplot(alpha = 0.3) +
    geom_jitter(size = 0.4, alpha = 0.4) +
    ggtitle("Bone Mineral Denisty vs Age at First Birth Before & At/After 20yo") + xlab("Age at First Birth (yrs)") + ylab("Femoral Neck BMD (g/cm^2)") + scale_fill_viridis("Race/Ethnicity", discrete = TRUE, option = "A")

# view the plot
print(lm_5)
```








