---
title: "analysis"
author: "Calli Quire"
date: "2025-02-14"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Set up
## Download, install, run relevant packages
```{r, warning=FALSE, message=FALSE, echo=FALSE}
# Install relevant packages
install.packages("dplyr")
install.packages("survey")
install.packages("haven")
install.packages("tidyr")
install.packages("ggplot2")
install.packages("ggbeeswarm")
install.packages("tidyverse")
install.packages("hrbrthemes")
install.packages("viridis")
install.packages("nlme")
install.packages("lme4")
install.packages("tseries")
install.packages("cowplot")
install.packages("sjPlot")
install.packages("sjmisc") 
install.packages("effects")
install.packages("sjstats")
install.packages("WriteXLS")
install.packages("AICcmodavg")
```

```{r echo=FALSE, message=TRUE, warning=FALSE, paged.print=TRUE}
# Load relevant packages
# + message = FALSE, warning=FALSE
library(dplyr)
library(survey)
library(haven)
library(workflowr)
library(tidyr)
library(ggplot2)
library(ggbeeswarm)
library(tidyverse)
library(hrbrthemes)
library(viridis)
library(nlme)
library(lme4)
library(tseries)
library(cowplot) 
library(sjPlot) 
library(sjmisc) 
library(effects)
library(sjstats)
library(WriteXLS)
library(AICcmodavg)
#'
options(survey.lonely.psu='adjust')

# Display Version Information
cat("R package versions:\n")
for (p in c("base", "survey","dplyr"))
{ cat(p, ": ", as.character(packageVersion(p)), "\n")}
```


# Import data
## Download and read data
### 1. Demographic data
```{r}
## Demographic (DEMO) data
## isolate participant identification/sequence number, gender, age, variances, race/Hispanic origin

# 2007-2008
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2007/DataFiles/DEMO_E.XPT", tf <- tempfile(), mode="wb")
DEMO_E <- foreign::read.xport(tf)[,c("SEQN","RIAGENDR","RIDAGEYR","SDMVSTRA","SDMVPSU","WTMEC2YR", "RIDRETH1")]

# 2009-2010
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2009/DataFiles/DEMO_F.xpt", tf <- tempfile(), mode="wb")
DEMO_F <- foreign::read.xport(tf)[,c("SEQN","RIAGENDR","RIDAGEYR","SDMVSTRA","SDMVPSU","WTMEC2YR", "RIDRETH1")]

# 2013-2014
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2013/DataFiles/DEMO_H.XPT", tf <- tempfile(), mode="wb")
DEMO_H <- foreign::read.xport(tf)[,c("SEQN","RIAGENDR","RIDAGEYR","SDMVSTRA","SDMVPSU","WTMEC2YR", "RIDRETH1")]

# 2017-2018
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/DEMO_J.XPT", tf <- tempfile(), mode="wb")
DEMO_J <- foreign::read.xport(tf)[,c("SEQN","RIAGENDR","RIDAGEYR","SDMVSTRA","SDMVPSU","WTMEC2YR", "RIDRETH1")]

```


### 2. Reproductive health question (RHQ) data
```{r}
# Reproductive Health Questions (RHQ)
# isolate sequence number, age at first live birth, has had regular periods in the last 12 mo, reason for not having regular periods in last 12 mo

# 2007-2008
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2007/DataFiles/RHQ_E.XPT", tf <- tempfile(), mode="wb")
RHQ_E <- foreign::read.xport(tf)[,c("SEQN","RHD180", "RHQ031", "RHD042", "RHQ171")]

# 2009-2010
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2009/DataFiles/RHQ_F.xpt", tf <- tempfile(), mode="wb")
RHQ_F <- foreign::read.xport(tf)[,c("SEQN","RHD180", "RHQ031", "RHD042", "RHQ171")]

# 2013-2014
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2013/DataFiles/RHQ_H.xpt", tf <- tempfile(), mode="wb")
RHQ_H <- foreign::read.xport(tf)[,c("SEQN","RHD180", "RHQ031", "RHD043", "RHQ171")]

# 2017-2018
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/RHQ_J.xpt", tf <- tempfile(), mode="wb")
RHQ_J <- foreign::read.xport(tf)[,c("SEQN","RHD180", "RHQ031", "RHD043", "RHQ171")]

```


### 3. Bone mineral denisty (BMD) data
```{r}
# Dual-Energy X-Ray Absorptiometry - Total Femur
# isolate sequence number, total femoral BMD

# 2007-2008
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2007/DataFiles/DXXFEM_E.xpt", tf <- tempfile(), mode="wb")
DXX_E <- foreign::read.xport(tf)[,c("SEQN", "DXXNKBMD")]

# 2009-2010
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2009/DataFiles/DXXFEM_F.xpt", tf <- tempfile(), mode="wb")
DXX_F <- foreign::read.xport(tf)[,c("SEQN", "DXXNKBMD")]

# 2013-2014
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2013/DataFiles/DXXFEM_H.xpt", tf <- tempfile(), mode="wb")
DXX_H <- foreign::read.xport(tf)[,c("SEQN", "DXXNKBMD")]

# 2017-2018
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/DXXFEM_J.xpt", tf <- tempfile(), mode="wb")
DXX_J <- foreign::read.xport(tf)[,c("SEQN", "DXXNKBMD")]

```


# Clean
## Combine the data sets
```{r}
# combine data sets for DEMO, RHQ, and DXX into three sub-sets
DEMO <- bind_rows(DEMO_E, DEMO_F, DEMO_H, DEMO_J)
RHQ <- bind_rows(RHQ_E, RHQ_F, RHQ_H, RHQ_J) 
DXX <- bind_rows(DXX_E, DXX_F, DXX_H, DXX_J)
```

## Merge subsets into one main data set
```{r}
# merge DEMO + RHQ
predata <- left_join(DEMO, RHQ, by = "SEQN")

# merge 'predata' = DEMO, RHQ, + DXX
data <- left_join(predata, DXX, by = "SEQN")
```

## Combine any single variables with 2+ codes into single columns 
```{r}
# combine any variables that have two different variable names (e.g., RHD042 = RHD043) into single columns

# combine RHD042, RHD043
data$RHQ04X <- coalesce(data$RHD042, data$RHD043)
```

## Simplify variable/column names
```{r}
data <-
   rename(data, 
     BMD = DXXNKBMD, AFB = RHD180, MENS = RHQ031,
     GENDR = RIAGENDR, AGE = RIDAGEYR, RETH = RIDRETH1, NMENS = RHQ04X, PAR = RHQ171
     )
```

## Omit males from data
```{r}
# remove male study participants from analysis data; responses to RIAGENDR coded 1
data <- subset(data, !(GENDR %in% c(1)))
```

## Omit any participants without BMD (DXXBMD), AFB (RHX180), period/year (RHQ03X) data
```{r}
# missing BMD data
data <- drop_na(data, "BMD")

# missing AFB data
data <- drop_na(data, "AFB")

# missing period/year data
data <- drop_na(data, "MENS")

# missing parity data
data <- drop_na(data, "PAR")
```

## Omit any participants who answered "don't know" (coded 999), or "refused" (coded as 777) to AFB
```{r}
data <- subset(data, !(AFB %in% c(777, 999)))
```

## Omit any participants who answered "don't know" (coded 9), or "refused" (coded as 7) to RHQ03X (Period in last 12 mo?)
```{r}
# removes participants who don't know if they have had a period in the last 12 months
data <- subset(data, !(MENS %in% c(7, 9)))
```

## Omit any participants who answered 5-9, 77, 99 to RHQ04X (Why no period/year?)
```{r}
# removes participants who aren't having periods b/c: hysterectomy (coded 3), periods usually irregular (coded 4), going/gone through menopause (coded 5 or 7), medical conditions/treatments (coded 6 or 8), other (coded 9), refused (coded 77), don't know (coded 99)
# there are multiple codes for some conditions due to coding inconsistencies between sampling years
# excludes data on individuals who may not be in 'good health'-- defined 'good health' in exclusion flowchart
data <- subset(data, !(NMENS %in% c(3, 4, 5, 6, 7, 8, 9, 77, 99)))
```

## Omit any participants who answered refused or don't know to parity
```{r}
data <- subset(data, !(PAR %in% c(77, 99)))
```

## Add column for AFB b4 18, from 18-27, from 28+
```{r}
# adds column called BINS where the following is coded: 
data <- data %>%
  mutate(BINS = case_when(
    AFB > 18  & AFB < 27 ~ "19-26",
    AFB < 19 ~ "</= 18", 
    AFB > 26 ~ ">/= 27",
  ))
```

## Replace race/ethnicity codes with 'definitions'
```{r}
data <- data %>% mutate(RETH = recode(RETH, '1' = 'Mexican American', '2' = 'Other Hispanic', '3' = 'Non-Hispanic White', '4' = 'Non-Hispanic Black', '5' = 'Other'))

# STILL NEED TO COLLAPSE 'OTHER' CATEGORIES INTO ONE
```

## Create simplified data frames for analyses
```{r}
# drop all columns that are not the combined columns, age, gender, race/origin or sequence number
# makes simplified data frame (named "df") for analyses purposes
# does not delete the full data frame (named "data")
df <- subset(data, select = -c(SDMVSTRA, SDMVPSU, WTMEC2YR, RHD042, RHD043))

```


# Analysis 
## Without parity

### Exploratory analysis
```{r}
# simple multivariate regression
lm1 <- lm(BMD ~ AFB + AGE, data = df)
summary(lm1)

# added other potential variables
lm2 <- lm(BMD ~ AFB + AGE + RETH, data = df)
summary(lm2)

# testing interaction effects btw AFB and AGE
lm3 <- lm(BMD ~ AFB * AGE + RETH, data = df)
summary(lm3)

# non-linear effect of AGE (as a polynomial term)
lm4 <- lm(BMD ~ AFB + AGE + I(AGE^2) + RETH, data = df)
summary(lm4)

# find best fit model 
anova(lm1, lm2, lm3, lm4)

# lm2 seems to be best fit, but stats arent great
```

### Plots
```{r}
# visualize BMD vs AGE in scatter plot 
plot1 <- 
  ggplot(df, aes(x = AGE, y = BMD)) + geom_smooth(method = "lm") + 
  geom_point(alpha = 0.5) + 
  scale_x_continuous(breaks = seq(0, 70, by = 5)) + labs(title = "Bone Mineral Density vs Age at Time of Scan", y = "Femoral Neck BMD (g/cm^2)", x = "Age at Time of Scan (yrs)")
print(plot1)

# visualize BMD vs AFB in scatter plot
plot2 <- 
  ggplot(df, aes(x = AFB, y = BMD)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggtitle("Bone Mineral Density vs Age at First Birth") +
  xlab("Age at First Birth (yrs)") +
  ylab("Femoral Neck BMD (g/cm^2)") +
  scale_x_continuous(breaks = seq(0, 42, by = 2))
print(plot2)

# visualize BMD vs binned AFB in box plot
plot3 <-
  ggplot(df, aes(x = BINS, y = BMD, fill = BINS)) +
  geom_boxplot() +
  labs(title = "Bone Mineral Density vs Binned Age at First Birth", x = "Age at First Birth (yrs)", y = "Femoral Neck BMD (g/cm^2)", fill = "AFB") +
  scale_x_discrete(limits = c("</= 18","19-26",">/= 27")) +
  scale_fill_brewer(palette = "Dark2", breaks=c('</= 18', '19-26', '>/= 27'))
print(plot3)

# visualize BMD vs binned AFB in violin plot (by reported ancestry)
plot4a <-
  ggplot(df, aes(x = BINS, y = BMD, fill = RETH)) +
  geom_violin() +
  geom_jitter(size = 0.5, alpha = 0.5) +
  labs(title = "Bone Mineral Density vs Binned Age at First Birth", x = "Age at First Birth (yrs)", y = "Femoral Neck BMD (g/cm^2)", fill = "Reported Ancestry") +
  scale_x_discrete(limits = c("</= 18","19-26",">/= 27")) +
  scale_fill_brewer(palette = "Dark2")
print(plot4a)

# visualize BMD vs binned AFB in box plot (by reported ancestry)
plot4b <-
  ggplot(df, aes(x = BINS, y = BMD, fill = RETH)) +
  geom_boxplot() +
  labs(title = "Bone Mineral Density vs Binned Age at First Birth", x = "Age at First Birth (yrs)", y = "Femoral Neck BMD (g/cm^2)", fill = "Reported Ancestry") +
  scale_x_discrete(limits = c("</= 18","19-26",">/= 27")) +
  scale_fill_brewer(palette = "Dark2")
print(plot4b)

# visualize BMD vs AGE (by binned AFB) in scatter plot
plot5 <-
  ggplot(df, aes(x = AGE, y = BMD, color = BINS)) + 
  geom_point(alpha = 0.7) +
  labs(title = "Bone Mineral Density vs Age at Time of Scan", x = "Age at Time of Scan (yrs)", y = "Femoral Neck BMD (g/cm^2)", color = "Age at First Birth (yrs)") +
  scale_color_brewer(palette = "Dark2", breaks=c('</= 18', '19-26', '>/= 27')) +
  scale_x_continuous(breaks = seq(0, 70, by = 5))
print(plot5)

```


# Clean
## With parity 

## Merge subsets into one main data set
```{r}
# first, need to clean with parity focus

# merge DEMO + RHQ
predata2 <- left_join(DEMO, RHQ, by = "SEQN")

# merge 'predata' = DEMO, RHQ, + DXX
data2 <- left_join(predata, DXX, by = "SEQN")
```

## Combine any single variables with 2+ codes into single columns
```{r}
# combine any variables that have two different variable names (e.g., RHD042 = RHD043) into single columns

# combine RHD042, RHD043
data2$RHQ04X <- coalesce(data2$RHD042, data2$RHD043)
```

## Simplify variable/column names
```{r}
data2 <-
   rename(data2, 
     BMD = DXXNKBMD, AFB = RHD180, MENS = RHQ031,
     GENDR = RIAGENDR, AGE = RIDAGEYR, RETH = RIDRETH1, NMENS = RHQ04X, PAR = RHQ171
     )
```

## Omit males from data
```{r}
# remove male study participants from analysis data; responses to RIAGENDR coded 1
data2 <- subset(data2, !(GENDR %in% c(1)))
```

## Omit any participants without BMD or PARITY data
```{r}
# missing BMD data
data2 <- drop_na(data2, "BMD")

# missing parity data
data2 <- drop_na(data2, "PAR")
```

## Omit any participants who answered "don't know" (coded 9), or "refused" (coded as 7) to RHQ03X (Period in last 12 mo?)
```{r}
# removes participants who don't know if they have had a period in the last 12 months
data2 <- subset(data2, !(MENS %in% c(7, 9)))
```

## Omit any participants who answered 5-9, 77, 99 to RHQ04X (Why no period/year?)
```{r}
# removes participants who aren't having periods b/c: hysterectomy (coded 3), periods usually irregular (coded 4), going/gone through menopause (coded 5 or 7), medical conditions/treatments (coded 6 or 8), other (coded 9), refused (coded 77), don't know (coded 99)
# there are multiple codes for some conditions due to coding inconsistencies between sampling years
# excludes data on individuals who may not be in 'good health'-- defined 'good health' in exclusion flowchart
data2 <- subset(data2, !(NMENS %in% c(3, 4, 5, 6, 7, 8, 9, 77, 99)))
```

## Omit any participants who answered refused or don't know to parity
```{r}
data2 <- subset(data2, !(PAR %in% c(77, 99)))
```

## Add column for AFB b4 18, from 18-27, from 28+
```{r}
# adds column called BINS where the following is coded: 
data2 <- data2 %>%
  mutate(BINS = case_when(
    AFB > 18  & AFB < 27 ~ "19-26",
    AFB < 19 ~ "</= 18", 
    AFB > 26 ~ ">/= 27",
  ))
```

## Replace race/ethnicity codes with 'definitions'
```{r}
data2 <- data2 %>% mutate(RETH = recode(RETH, '1' = 'Mexican American', '2' = 'Other Hispanic', '3' = 'Non-Hispanic White', '4' = 'Non-Hispanic Black', '5' = 'Other'))

# STILL NEED TO COLLAPSE 'OTHER' CATEGORIES INTO ONE
```

## Replace parity codes with 'definitions'
```{r}
data2 <- data2 %>% mutate(PAR = recode(PAR, '0' = 'Nulliparous', '1' = '1', '2' = '2', '3' = '3', '4' = '4', '5' = '5', '6' = '6', '7' = '7', '8' = '8', '9' = '9', '10' = '10+'))
```

## Create simplified data frames for analyses
```{r}
# drop all columns that are not the combined columns, age, gender, race/origin or sequence number
# makes simplified data frame (named "df") for analyses purposes
# does not delete the full data frame (named "data")
df2 <- subset(data2, select = -c(SDMVSTRA, SDMVPSU, WTMEC2YR, RHD042, RHD043))
```

# Analysis 
## With parity

### Exploratory analysis
```{r}
lm5 <- 
  lm(BMD ~ PAR, data = df2)
summary(lm5)

lm6 <- 
  lm(BMD ~ PAR + AGE, data = df2)
summary(lm6)

lm7 <-
  lm(BMD ~ PAR + AGE, data = df2)
summary(lm7)

lm8 <-
  lm(BMD ~ PAR + AGE + RETH, data = df2)
summary(lm8)

lm9 <-
  lm(BMD ~ PAR * AGE + RETH, data = df2)
summary(lm9)

lm10 <-
  lm(BMD ~ PAR + I(AGE^2) + RETH, data = df2)
summary(lm10)

# find best fit model 
anova(lm5, lm6, lm7, lm8, lm9, lm10)

# lm8, lm10 seem best fits
```

# Plots
```{r}
# visualize BMD vs PAR (by age at time of scan) in scatter plot
plot6 <- 
  ggplot(df2, aes(x = PAR, y = BMD, color = AGE)) +
  geom_point() +
  labs(title = "Bone Mineral Density vs Parity in Premenopausal Participants", x = "Parity (live births/person)", y = "Femoral Neck BMD (g/cm^2)", color = "Age at Time of Scan") +
 scale_x_discrete(limits = c("Nulliparous","1","2", "3", "4", "5", "6", "7", "8", "9", "10+")) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  scale_color_viridis()
print(plot6)


# Binned BMD v Parity
## avg BMD for femoral neck ranges from 0.68 to 0.95 depending on age at time of scan and reported ancestry

## make new column for BMD bins called BMDCAT:
df2 <- df2 %>%
  mutate(BMDCAT = case_when(
    BMD > 0.68 & BMD < 0.95 ~ "Average",
    BMD < 0.69 ~ "Below Average", 
    BMD > 0.94 ~ "Above Average"
  ))

## make new column for AGE bins called AGECAT:
df2 <- df2 %>%
  mutate(AGECAT = case_when(
    AGE > 19 & AGE < 30 ~ "20-29",
    AGE > 29 & AGE < 40 ~ "30-39", 
    AGE > 39 & AGE < 50 ~ "40-49",
    AGE > 49 & AGE < 60 ~ "50-59",
    AGE > 59 & AGE < 70 ~ "60-69"
  ))

# make table with sample sizes to add (n =) into bar chart
parity_count <- df2 %>% 
  group_by(PAR) %>% 
  summarise(number_cases = n())
print(parity_count)

## plot frequency of below-avg BMD v parity groups
plot7 <-
  ggplot(df2, aes(fill = BMDCAT, y = BMD, x = PAR)) + 
  geom_bar(position = "fill", stat = "identity") +
  labs(title = "Frequency of Below Average BMD vs Parity", x = "Parity (live births/participant)", y = "% of Participants", fill = "BMD (Femoral Neck)") +
  scale_x_discrete(limits = c("Nulliparous","1","2", "3", "4", "5", "6", "7", "8", "9", "10+")) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  scale_fill_brewer(palette = "Dark2")
print(plot7)

## plot frequency of below-avg BMD v Parity groups BY AGE category 

## one plot for each age cohort: 20s
plot20s <-
  df2 %>%
  filter(AGECAT == c("20-29")) %>%
  ggplot(aes(fill = BMDCAT, y = BMD, x = PAR)) + 
  geom_bar(position = "fill", stat = "identity") +
  labs(title = "Frequency of Below Average BMD vs Parity for participants aged 20-29", x = "Parity (live births/participant)", y = "% of Participants", fill = "BMD (Femoral Neck)") +
  scale_x_discrete(limits = c("Nulliparous","1","2", "3", "4", "5")) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  scale_fill_brewer(palette = "Dark2")
print(plot20s)

## one plot for each age cohort: 30s
plot30s <-
  df2 %>%
  filter(AGECAT == c("30-39")) %>%
  ggplot(aes(fill = BMDCAT, y = BMD, x = PAR)) + 
  geom_bar(position = "fill", stat = "identity") +
  labs(title = "Frequency of Below Average BMD vs Parity for participants aged 30-39", x = "Parity (live births/participant)", y = "% of Participants", fill = "BMD (Femoral Neck)") +
  scale_x_discrete(limits = c("Nulliparous","1","2", "3", "4", "5", "6", "7", "8")) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  scale_fill_brewer(palette = "Dark2")
print(plot30s)

## one plot for each age cohort: 40s
plot40s <-
  df2 %>%
  filter(AGECAT == c("40-49")) %>%
  ggplot(aes(fill = BMDCAT, y = BMD, x = PAR)) + 
  geom_bar(position = "fill", stat = "identity") +
  labs(title = "Frequency of Below Average BMD vs Parity for participants aged 40-49", x = "Parity (live births/participant)", y = "% of Participants", fill = "BMD (Femoral Neck)") +
  scale_x_discrete(limits = c("Nulliparous","1","2", "3", "4", "5", "6", "7", "8", "9", "10+")) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  scale_fill_brewer(palette = "Dark2")
print(plot40s)

## one plot for each age cohort: 50s
plot50s <-
  df2 %>%
  filter(AGECAT == c("50-59")) %>%
  ggplot(aes(fill = BMDCAT, y = BMD, x = PAR)) + 
  geom_bar(position = "fill", stat = "identity") +
  labs(title = "Frequency of Below Average BMD vs Parity for participants aged 50-59", x = "Parity (live births/participant)", y = "% of Participants", fill = "BMD (Femoral Neck)") +
  scale_x_discrete(limits = c("Nulliparous","1","2", "3", "4", "5", "6", "7", "8", "9", "10+")) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  scale_fill_brewer(palette = "Dark2")
print(plot50s)

## one plot for each age cohort: 60s
plot60s <-
  df2 %>%
  filter(AGECAT == c("60-69")) %>%
  ggplot(aes(fill = BMDCAT, y = BMD, x = PAR)) + 
  geom_bar(position = "fill", stat = "identity") +
  labs(title = "Frequency of Below Average BMD vs Parity for participants aged 60-69", x = "Parity (live births/participant)", y = "% of Participants", fill = "BMD (Femoral Neck)") +
  scale_x_discrete(limits = c("1","2", "3", "4", "5", "6")) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  scale_fill_brewer(palette = "Dark2")
print(plot60s)
```

