---
title: "nhanes_exclusion_cleaning"
author: "cquire"
date: "2025-01-26"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Cleaning

## Downloading and Importing Datafiles
```{r, warning=FALSE, message=FALSE, echo=FALSE}
# Install relevant packages
install.packages("dplyr")
install.packages("survey")
install.packages("haven")
install.packages("tidyr")
install.packages("ggplot2")
install.packages("ggbeeswarm")
```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
#Load relevant packages
#+ message = FALSE, warning=FALSE
library(dplyr)
library(survey)
library(haven)
library(workflowr)
library(tidyr)
library(ggplot2)
library(ggbeeswarm)
#'
options(survey.lonely.psu='adjust')

# Display Version Information
cat("R package versions:\n")
for (p in c("base", "survey","dplyr"))
{ cat(p, ": ", as.character(packageVersion(p)), "\n")}
```

```{r}
# Download and read data

# Demographic (DEMO)
# isolated participant identification/sequence number, gender, age, variances
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2007/DataFiles/DEMO_E.XPT", tf <- tempfile(), mode="wb")
DEMO_E <- foreign::read.xport(tf)[,c("SEQN","RIAGENDR","RIDAGEYR","SDMVSTRA","SDMVPSU","WTMEC2YR")]
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2009/DataFiles/DEMO_F.xpt", tf <- tempfile(), mode="wb")
DEMO_F <- foreign::read.xport(tf)[,c("SEQN","RIAGENDR","RIDAGEYR","SDMVSTRA","SDMVPSU","WTMEC2YR")]

# Reproductive Health Questions (RHQ)
# isolated sequence number, age at first live birth, has had regular periods in the last 12 mo, reason for not having regular periods in last 12 mo
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2007/DataFiles/RHQ_E.XPT", tf <- tempfile(), mode="wb")
RHQ_E <- foreign::read.xport(tf)[,c("SEQN","RHD180", "RHQ031", "RHD042")]
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2009/DataFiles/RHQ_F.xpt", tf <- tempfile(), mode="wb")
RHQ_F <- foreign::read.xport(tf)[,c("SEQN","RHD180", "RHQ031", "RHD042")]

# Dual-Energy X-ray Absorptiometry - Total Spine (DXXSPN)
# isolated sequence number, total spine BMD
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2007/DataFiles/DXXSPN_E.xpt", tf <- tempfile(), mode="wb")
DXXSPN_E <- foreign::read.xport(tf)[,c("SEQN", "DXXOSBMD")]
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2009/DataFiles/DXXSPN_F.xpt", tf <- tempfile(), mode="wb")
DXXSPN_F <- foreign::read.xport(tf)[,c("SEQN", "DXXOSBMD")]

# Dual-Energy X-ray Absorptiometry - Femur (DXXFEM)
# isolated sequence number, femoral neck BMD
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2007/DataFiles/DXXFEM_E.xpt", tf <- tempfile(), mode="wb")
DXXFEM_E <- foreign::read.xport(tf)[,c("SEQN", "DXXNKBMD")]
download.file("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2009/DataFiles/DXXFEM_F.xpt", tf <- tempfile(), mode="wb")
DXXFEM_F <- foreign::read.xport(tf)[,c("SEQN", "DXXNKBMD")]

```

## Data Cleaning
```{r}
# combining D, E data sets
DEMO <- bind_rows(DEMO_E, DEMO_F)
RHQ <- bind_rows(RHQ_E, RHQ_F) 
DXXSPN <- bind_rows(DXXSPN_E, DXXSPN_F)
DXXFEM <- bind_rows(DXXFEM_E, DXXFEM_F)
```

## Merge DEMO, RHQ, DXX, DXXFEM data sets
```{r}
# merging DEMO and RHQ
data <- left_join(DEMO, RHQ, by = "SEQN")

# merging data (includes DEMO, RHQ) w/ DXXSPN
medium_data <- left_join(data, DXXSPN, by = "SEQN")

# merging medium_data (includes DEMO, RHQ, DXXSPN) w/ DXXFEM
mega_data <- left_join(medium_data, DXXFEM, by = "SEQN")
```

## Omit missing (N/A) AFB RHQ values
```{r}
# removing missing AFB values from data set; will subsequently remove all males from set
fem_mega_data <- drop_na(mega_data, "RHD180")
```

## Remove menopausal and postmenopausal participants
*** want to omit participants who are menopausal or postmenopausal; can do so by omitting those who are:
(1) not having a menstrual period in past 12mo and not b/c they were pregnant, breastfeeding, or had a hysterectomy, and
  (1A) if had hysterectomy, are they below the avg age of menopause onset 49 yo

```{r}
# omit missing (N/A) values from RHQ031
fem_mega_data_1 <- drop_na(fem_mega_data, "RHQ031")

# omit participants who answer menopause (7), other (9), refused (77), don't know (99) to RHD043
# this removes participants who aren't having periods b/c of menopause, another reason, or refused/don't know
fem_mega_data_2 <- subset(fem_mega_data_1, !(RHD042 %in% c(7, 8, 9, 77, 99)))

# omit participants who answer >/= 49 to RIDAGEYR
# this removes participants who may have had a hysterectomy, and therefore cannot say if they are experiencing menopause (based on lack of menstruation), and are at or above the average age that menopause occurs  
fert_fem_mega_data <- subset(fem_mega_data_2, !(RIDAGEYR %in% c(49, 50, 51, 52, 53, 54, 55, 56, 57, 58, 59, 60, 61, 62, 63, 64, 65, 66, 67, 68, 69, 70, 71, 72, 73, 74, 75, 76, 77, 78, 79, 80)))
```

## Omit participants without BMD data
```{r}
# this removes any participants who are missing BMD data
bmd_fert_fem_mega_data <- drop_na(fert_fem_mega_data, "DXXOSBMD", "DXXNKBMD")
```

## Add PBMDA column for AFB before 20yo vs at/after 20yo
```{r}
# this adds a column for if participants' age at first birth was before age of peak bone mineral density accrual (PBMDA) or at/after age of PBMDA
# if birth occurred before participant was 20, coded 1; if occurred on or after participant was 20, coded 0
bmd_fert_fem_mega_data <- bmd_fert_fem_mega_data %>%
  mutate(PBMDA = ifelse(RHD180 < 20, "yes", "no" ))
```


# Analysis

## Univariate Linear Regressions

### Run univariate linear regression with Whole Spine BMD (DXXOSBMD) vs AFB (RHD180) 
```{r}
# run an univariate linear regression where independent/x = AFB, dependent/y = whole spine BMD
ulm_os <- lm(DXXOSBMD ~ RHD180, data = bmd_fert_fem_mega_data)

# visualize ulm_os via scatter plot
ulm_os_plot <- ggplot(ulm_os, aes(x = RHD180, 
            y = DXXOSBMD)) + geom_point(alpha = 0.5) + scale_x_continuous("AFB (yrs)", breaks = seq(0, 40, by = 2)) + scale_y_continuous("Whole Spine Bone Mineral Density (g/cm^2)", breaks = seq(0, 2, by = 0.1)) + annotate("rect", xmin = -Inf, xmax = Inf, ymin = 1.1, ymax = 1.2, fill = "lightgreen", alpha = 0.2) + geom_vline(xintercept = 20, linetype = "dashed", color = "darkorange") + geom_vline(xintercept = 22, linetype = "dashed", color = "yellow") + geom_vline(xintercept = 18, linetype = "dashed", color = "yellow")
  
# view the plot
print(ulm_os_plot)
```

### Run univariate linear regression with Femoral Neck BMD (DXXLSBMD) vs AFB (RHD180)
```{r}
# run an univariate linear regression where independent/x = AFB, dependent/y = femoral neck BMD
ulm_nk <- lm(DXXNKBMD ~ RHD180, data = bmd_fert_fem_mega_data)

# visualize ulm_nk via scatter plot
# ***want to add green rectangle for where avg femoral neck BMD sits***
ulm_nk_plot <- ggplot(ulm_nk, aes(x = RHD180, 
            y = DXXNKBMD)) + geom_point(alpha = 0.5) + scale_x_continuous("AFB (yrs)", breaks = seq(0, 40, by = 2)) + scale_y_continuous("Femoral Neck Bone Mineral Density (g/cm^2)", breaks = seq(0, 2, by = 0.1)) + geom_vline(xintercept = 20, linetype = "dashed", color = "darkorange") + geom_vline(xintercept = 22, linetype = "dashed", color = "yellow") + geom_vline(xintercept = 18, linetype = "dashed", color = "yellow")
  
# view the plot
print(ulm_nk_plot)
```

### Run univariate linear regression with Whole Spine BMD (DXXOSBMD) vs binary AFB (PBMDA)
```{r}
# run an univariate linear regression where x = AFB (binary), y = whole spine BMD
ulm_os_bin <- lm(DXXOSBMD ~ PBMDA, data = bmd_fert_fem_mega_data)

# visualize ulm_os_bin via beeswarm plot
ulm_os_bin_plot <- ggplot(ulm_os_bin, aes(x = PBMDA, y = DXXOSBMD)) +
  geom_beeswarm() + scale_x_continuous("AFB before OR on/after PBMDA (no/yes)", breaks=seq(0,1,by=1)) + scale_y_continuous("Whole Spine Bone Mineral Density (g/cm^2)", breaks = seq(0, 2, by = 0.1)) + annotate("rect", xmin = -Inf, xmax = Inf, ymin = 1.1, ymax = 1.2, fill = "lightgreen", alpha = 0.2)

# view the plot
print(ulm_os_bin_plot)
```

### Run univariate linear regression with Femoral Neck BMD (DXXOSBMD) vs binary AFB (PBMDA)
```{r}
# run an univariate linear regression where x = AFB (binary), y = femoral neck BMD
ulm_nk_bin <- lm(DXXNKBMD ~ PBMDA, data = bmd_fert_fem_mega_data)

# visualize ulm_nk_bin via beeswarm plot
# ***want to add green rectangle for where avg femoral neck BMD sits***
ulm_nk_bin_plot <- ggplot(ulm_nk_bin, aes(x = PBMDA, y = DXXNKBMD)) +
  geom_beeswarm() + scale_x_continuous("AFB before OR on/after PBMDA (no/yes)", breaks=seq(0,1,by=1)) + scale_y_continuous("Femoral Neck Bone Mineral Density (g/cm^2)", breaks = seq(0, 2, by = 0.1))

# view the plot
print(ulm_nk_bin_plot)
```

## Multivariate/Multiple Linear Regressions

### Run multiple linear regression with Whole Spine BMD (DXXOSBMD) vs AFB (RHD180) + binary AFB (PBMDA)
```{r}
# run analysis
mlm_os <- lm(DXXOSBMD ~ RHD180 + PBMDA, data = bmd_fert_fem_mega_data)
summary(mlm_os)
confint(mlm_os)

# visualize
mlm_os_plot <- ggplot(mlm_os, aes(y = DXXOSBMD, x = RHD180, color = PBMDA)) + geom_point() + scale_x_continuous("AFB (yrs)", breaks = seq(0, 40, by = 2)) + scale_y_continuous("Whole Spine Bone Mineral Density (g/cm^2)", breaks = seq(0, 2, by = 0.1)) + scale_color_continuous("AFB before OR on/after PBMDA (no/yes)", breaks = seq(0, 1, by = 1))

# view 
print(mlm_os_plot)
```

### Run multivariate linear regression with Femoral Neck BMD (DXXNKBMD) vs AFB (RHD180) + binary AFB (PBMDA)
```{r}
mlm_nk <- lm(DXXNKBMD ~ RHD180 + PBMDA, data = bmd_fert_fem_mega_data)
summary(mlm_nk)
confint(mlm_nk)
```

